Application for Rethinking Foundations of Physics 2017

### 2017-01-14 Rethinking Foundations of Physics 2017/application.tex
application.tex \subsection{Discussion} Added, and cleaned up...

If we insist that probabilities cannot be computed to infinite precision
and are bound to be approximations represented by intervals of confidence,
then quantum states themselves can only be discussed within intervals
of confidence. This raises an important philosophical question with
immense practical relevance in the context of quantum computing. The
question, which we could largely bypass in the classical case, is
whether there is, in the background of all the measurements and probabilities,
a ``real'' infinitely-precise quantum state that is being approximated
and that, in principle, can be more and more accurately described
given more and more resources.

@@

\subsection{Quantum Bayesianism}

Although we discussed how a quantum IVPM could be induced from a real
quantum state, our investigation leaves an open question of whether
there exists a ``real'' entity that exists independently of measurements
and probabilities. Since the frequency interpretation of probability
is problematic, it is much more sensible to interpret probability
in the Bayesian sense as ``lack of knowledge''. This, and similar
arguments elegantly expressed by recent work on Quantum Bayesianism~\cite{Fuchs2010,VonBaeyer2016,Fuchs2012},
suggest that the quantum state is more like an interactive system
in computer science parlance. In other words, the state is only determined
by its interactions with the environment or observers. A rough but
useful analogy in the context of computer science would be a web server:
the state of such a system cannot be said to exist independently of
the interactions with clients; each interaction with a client determines
a future state of the system and may affect future interactions. Crucially,
each client may have a \emph{different} view of the web server that
is consistent with their own history of interactions. This computational
perspective agrees with the philosophical position of QBism~\cite{VonBaeyer2016}
that asserts that the quantum state is subjective: each observer has
a different view of the quantum system that is consistent with their
previous observations and that allows that observer, independently
of other observers, to assign beliefs (i.e., probabilities) to possible
future interactions.

\section{Kick-off Talk: The Reality of the Quantum State}

The meaning of quantum probability

The reality of the quantum state

— next section

The meaning of quantum probability: probability is approximate; intervals;
Born; no state when approximation is bad.

quantum state: Bayesian perspective; probability is subjective; hence
state is subjective; given previous results that show that state does
not exist when you have poor approximations; ?

In contrast to our previous works that assume the current quantum
theory is just an approximation to the discrete reality, we now assume
the nature really use real numbers. However, these real numbers could
not be directly accessed by human who is bounded by finite resources.
Therefore, even if the nature might have more computational power
than a Turing machine, human might not be able to acquire and use
the computational power more than a Turing machine.

In order to model the bounded-resource human observers, we would like
the theories of measurement and probability replaced by a subjective
probability model, called the interval-valued probability. The question
is how quantum mechanics would like with the interval-valued probability.

In particular, based on conventional probability, Gleason's theorem~\cite{gleason1957,Redhead1987-REDINA,peres1995quantum}
suggests that idea to define a state as a density matrix~\cite{Varadarajan2008}.
Since Gleason's theorem could not directly apply on interval-valued
probability measures, we would like to ask: How a quantum state looks
like with the interval-valued probability? When the number and precision
of the intervals increase, would the quantum interval-valued probability
measures converge to the quantum probability measure induced by the
Born rule~\cite{Born1983,Mermin2007,Jaeger2007}?

In previous work, we examined the Hilbert space mathematical framework
of quantum mechanics and teased apart its various aspects that directly
rely on the infinite-precision real or complex numbers. We note two
remarkable earlier results and an open question: 
\begin{itemize}
\item UNIQUE-SAT: Schumacher and
  Westmoreland~\cite{Schumacher2012-SCHMQT} argue that much of the
  structure of traditional quantum mechanics is maintained in the
  presence of finite fields. In particular, they establish that the
  quantum theory based on the finite field of booleans retains the
  following characteristics of quantum mechanics: the notions of
  superposition, interference, entanglement, and mixed states of
  quantum systems; the time evolution of quantum systems using
  invertible linear operators; the complementarity of incompatible
  observables; the exclusion of local hidden variable theories and the
  impossibility of cloning quantum states; and the presence of natural
  counterparts of quantum information protocols such as superdense
  coding and teleportation. However we have argued that this theory is
  not physically realizable as it allows the solution of
  computationally difficult problems like
  UNIQUE-SAT~\cite{Valiant198685} in constant time~\cite{usat}.

\item Deutsch-Jozsa: In a follow-up paper~\cite{DQT2014}, we proposed
  a variant of quantum mechanics in which the underlying field of
  complex numbers is replaced by special finite fields with enough
  structure to model Hermitian products that \emph{locally} behave
  like inner products. The trick is to recognize that, as the
  characteristic of the finite field increases, larger and larger
  regions of the finite field become totally ordered, thereby
  restoring, with some restrictions, all the usual properties of the
  inner product needed for conventional quantum mechanics. Using this
  theory to analyze the Deutsch-Jozsa algorithm reveals that the
  characteristic of the finite field plays an essential role: as the
  size of the finite field grows, so does the numerical range of the
  intermediate and final results of the algorithms being implemented.
  The theory naturally recovers a deterministic measure of the
  intrinsic resources required for a given level of complexity; this
  measure is normally completely hidden in computations with real
  numbers, and explicitly exposing it is one of the significant
  achievements of our discrete field analysis of quantum computation.
  This solves the conundrum that the conventional Deutsch-Jozsa
  algorithm~\cite{DeutschJozsa1992,Jaeger2007}
  mysteriously continues to work for larger and larger input
  functions without any apparent increase in resources.  Our analysis
  of this problem reveals that as the size of the input increases, it
  is necessary to increase the size of the finite field and hence the
  size of the underlying available numeric coefficients.  This
  observation does not fully explain the power of quantum computing
  over classical computing, but at least it explains that some of the
  power of quantum computing depends on increasingly larger precision
  in the underlying field of numbers.

\item An important special case of a generalized measurement is the
  symmetric, informationally complete, positive operator valued
  measure (SIC-POVM), which has applications in quantum
  cryptography. This problem has been formulated in the context of
  standard quantum theory where fields are complex numbers.  Although
  a rigorous proof for the existence of SIC-POVMs for arbitrary
  Hilbert space dimensions is still missing, it would be illuminating
  to look for a formulation of the same problem in a discrete quantum
  theory defined in terms of finite Galois fields.
\end{itemize}

A long term, ultimate, goal of this research proposal would be to
combine the various pieces into a coherent computational model of
quantum mechanics that keeps explicit track of resources needed to
represent the interactive system representing the quantum state, its
evolution, the observables, and the probabilities. Such a
computational model of quantum mechanics would encompass classical
(reversible) computation and potentially revolutionize quantum
thinking with finite resources. The explicit manipulation of resources
enables one of key idea that emerged in our previous work: the
resources used to represent the quantum system are in principle
unrelated to the resources available to the various observers. We
conjecture that this asymmetry may prove key to resolving some of the
quantum paradoxes.  The nature of quantum mechanics realized in terms
of discrete fields and finitely-communicable probability estimates is
amenable to direct computational exploration for reasonable field
sizes and observational accuracy. We have established a set of utility
libraries in Mathematica that supports a wide variety of such
numerical experiments. Concurrently to this effort, we are also
building on our previous work~\cite
{VDS2009,AGVS2005,VAS2006,finiteQC,Sabry:2003:MQC:871895.871900}
to build semantically well-founded implementations of the
computational models investigated in this proposal.

\paragraph*{Research Questions.} The aim of these implementations
include:
\begin{itemize}
\item Provide an experimental testbed to simulate physical phenomena
  (e.g.~\cite {OrtizEtAl2001,SommEtAl2002}) that complements the
  Mathematica libraries in which quantum algorithms can be simulated;
\item Establish connections between quantum programming concepts and
  classical programming concepts. We plan to make our software
  available for educational and research purposes; and 
\item Analyze the computational complexity of quantum theories and
  simulations based on discrete fields. What other (apart from
  UNIQUE-SAT) supernatural algorithms can be implemented in
  computational models based on discrete fields?
\item Determine, through numerical computations, solutions for
  SIC-POVM for small vector space dimensions.
\end{itemize}

propose to revisit quantum mechanics, quantum information, quantum
computation, and the foundations of physics from this resource-aware
perspective.

Soon after the discovery of quantum effects in the early 1900s, it
became evident that the quantum world requires a fresh worldview, not
merely an extension of the repertoire of entities and constructions
which explain classical physics. In the past thirty years, however,
the first quantum revolution mostly relied on just that: extensions of
classical logics, classical models of computation, and classical
analysis of algorithms. Fundamentally, the foundations of logic and
computing still embrace, to a large extent, the notion that the truth
of a statement is ``absolute'' and independent of any reasoning,
understanding, or action, i.e., logical statements are viewed as
abstract entities for which it makes no sense to talk about
entanglement with an observer: their truth or falsity is absolute and
does not depend on the notion of observation. This absolute view of
truth and the use of classical logic for reasoning about this absolute
truth have both been critiqued for centuries by philosophers and other
scientists but it is the advent of quantum science that really
devastates this worldview of classical determinism. Indeed, quantum
science suggests that the properties of physical systems can only be
learned by observers interacting with the system, and that this
interaction may entangle the observer with the system and disturb the
properties of interest. This suggests that the possibility structure
of our universe is not a fixed, Boolean structure, as we supposed
classically, but is in fact a non-Boolean structure that changes
dynamically~\cite{bub2006}.

We believe that the foundations of physics should be revisited from a
logical and computational approach

More directly, world-renowned logician Girard states:
Our vision for this revised computational and logic theory builds upon
programming abstractions which assume evolving, incomplete, and
inconsistent data, and principles for reasoning about programs which
are based on finitely communicable evidence between subsystems or
systems and observers, rather than absolute truth. The most natural
way to achieve this shift in our ``computational thinking'' and to
consolidate many of the new promising trends in computational,
informational, and logical theory is to design new programming
models. For it is indeed a widely accepted thesis that programming
languages shape thought processes and design
solutions~\cite{iverson80,johansen-berg01}. Our approach therefore is
to develop new programming models based on both logical foundations
and pragmatic considerations, to connect them to current
computational, informational, and logical theory, and to make
compelling arguments for the models by using them to write several
novel and challenging applications.

There is significant IU expertise in logic foundations, including
unconventional logics such paraconsistent logics, relevance logic,
multivalued logic with Bayesian inference, situation theory, and the
Dempster-Shafer theory of belief functions and plausible reasoning,
which can handle incomplete, evolving, and inconsistent information in
a controlled and discriminating way. There is also significant IU
expertise in the design and implementation of programming models, type
theories, and semantics that encompass both the classical and quantum
worlds. What IU lacks in expertise is an expert that bridges the
programming models with actual quantum algorithms: this is a new breed
of theoreticians with a strong background in both physics, complexity
theory, and cryptography.

Our initial results in that domain showed how subtle the
issues can be~\cite{usat,geometry2013,DQT2014}: a straightforward
replacement of the complex numbers by a finite field yields a variant
of quantum mechanics in which computationally hard problems like
UNIQUE-SAT (which decides whether a given Boolean formula is
unsatisfiable or has exactly one satisfying assignment) can be
deterministically solved in constant time. To eliminate such
unrealistic theories requires delicate analysis of the structure of
the Hilbert space, the process of observation, and the notion of
probability teasing apart their reliance on the infinitely precise
real numbers~\cite{geometry2013,DQT2014}. In this proposal, our aim is
to shift focus from the infinitely-specified but not directly
observable quantum states, to observable measurable properties of
quantum systems and their probabilities. Furthermore, we insist that
our theories of measurement and probability only refer to finitely
communicable evidence within feasible computational bounds. It follows
that states, observations, and probabilities all become ``fuzzy'',
i.e., specified by intervals of confidence that can only increase in
precision if the available resources increase proportionally. Our
notion of ``fuzzy quantum mechanics'' is related to existing
work~\cite{GranikCaulfield1996,Pykacz2013,SNL2009,Gudder2005,aerts1993physical}
but, as will be explained in more detail, is distinguished by its
unique computational character.

We will begin by reviewing existing work that recasts classical
probability spaces in a resource-aware setting and move to our
proposal which, briefly speaking, aims at recasting quantum
probability and hence quantum measurement to a corresponding
resource-aware setting.

In particular, the objectives to be achieved include:
\begin{itemize}
\item A computational framework for quantum measurement where quantum
  systems are described in terms of interactive processes. The
  framework takes care of the resources required (including the ones
  related to the observer) to determine a given property of the
  measured system.
\item Develop a measurement framework based on quantum interval-valued
  probabilities measures.
\item Assess the validity of fundamental theorems of quantum
  mechanics, such as Gleason, Bell, and Kochen-Specker, in a quantum
  formulation stripped away of infinite resources. We will aim to
  provide a definite answer to the Meyer-Mermin
  debate~\cite{PhysRevLett.83.3751, Mermin1999, BarrettKent2004} on
  the impact of finite precision measurements on the relevance of the
  Kochen-Specker
  theorem~\cite{kochenspecker1967,Redhead1987-REDINA,peres1995quantum,Jaeger2007}.
\item Study the computational complexity of quantum theories and
  simulations based on discrete (computable) fields.
\item Study the problem of finding a complete set of mutually unbiased
  bases
  (SIC-POVM)~\cite{1751-8121-45-8-085306,1751-8121-47-33-335302,Caves2002}
  over finite Galois fields.
\end{itemize}
We argue that no matter what the results are their impact will be
strong. Positive results that formulate a computable, with clear
resource bounds, theory of quantum measurement and quantum probability
would be essential for understanding and realizing quantum
computation. Any negative results would redirect research that aims at
realizing quantum computation to other approaches. At an intellectual
level, the results have the potential to solve or clarify several
paradoxes in quantum mechanics, quantum information, computability,
complexity theory, and perhaps quantum gravity.

For centuries, the mathematical formalization of physical processes
has been founded on the infinitely precise real or complex
numbers~\cite{Ziegler2007,weihrauch2012computable,blum2012complexity}.
For example, almost every description of quantum mechanics, quantum
computation, or quantum experiments refers to entities such as $\rme$,
$\pi$, $\sqrt{2}$, etc (see, e.g.,
\cite{Redhead1987-REDINA,544199,Mermin2007}).  Whether these models
based on real numbers faithfully describe the nature or just an
approximation is still an open question. If they are faithful, how the
nature store and compute these real numbers to produce our
experimental results would be the next natural question.  There are at
least three possibilities:
\begin{itemize}
\item For any computable number~\cite{Turing_1937}, the nature only store
the code of the Turing machine which computes the given number, and
manipulate these codes directly during the computation~\cite{Ziegler2007,weihrauch2012computable}.
\item The nature stores real numbers on infinitely long Turing machine tapes,
and handle these numbers by a Turing machine~\cite{Ziegler2007,weihrauch2012computable}.
In another word, the nature stores real numbers as a whole, but compute
them piecewisely.
\item The nature stores and computes real numbers as a whole~\cite{Ziegler2007,blum2012complexity}.
\end{itemize}
All these possibilities are not very desirable because they suggest
the nature might have more computational power than a Turing machine.
On one hand, if the nature only stores computable numbers by their
Turing machine codes, the nature might be able to decide whether two
Turing machines output exactly the same numbers, since this is used
to decide whether an observable is degenerated or not. However, testing
whether two Turing machines are output-equivalent is undecidable by
Rice's theorem~\cite{Rice_1953,Ziegler2007,weihrauch2012computable},
i.e., the nature has more computational power than a Turing machine.
On the other hand, if the nature stores real numbers as a whole, any
particular real number could serve as an oracle and give the nature
more computational power than a Turing machine. For example, maybe
the mass of the electron in some representation is the halting oracle
for some representation of Turing machines. Then, we could solve the
halting problem by just measuring the mass of the electron. This kind
of possibility has been online discussed by Peter Shor and others~\cite{PhysicsStackExchange16889}.

Since all possibilities that quantum theory is faithful are not very
desirable, maybe every description of quantum theory is just an approximation.
In this case, there is the other possibility:
\begin{itemize}
\item The real and complex numbers only exist in our model. The nature is
essentially discrete, and could be simulated by a Turing machine.
\end{itemize}
We have followed this approach to study quantum theory over discrete
finite fields~\cite{usat,geometry2013,DQT2014}. This yields a variant
of quantum mechanics in which computationally hard problems like UNIQUE-SAT
(which decides whether a given Boolean formula is unsatisfiable or
has exactly one satisfying assignment~\cite{Valiant198685,Papadimitriou1993,AroraBarak2009})
can be deterministically solved in constant time. We then delicate
analysis of the structure of the Hilbert space, the process of observation,
and the notion of probability by teasing apart their reliance on the
infinitely precise real numbers~\cite{geometry2013,DQT2014}.

>>>>

If we insist that probabilities cannot be computed to infinite precision
and are bound to be approximations represented by intervals of confidence,
then quantum states themselves can only be discussed within intervals
of confidence. This raises an important philosophical question with
immense practical relevance in the context of quantum computing. The
question is whether there is a ``real'' infinitely-precise quantum
state that is being approximated and that can be more and more accurately
described given more and more resources.

@@

\subsection{Discussion}

Although a family of more and more precise quantum IVPMs might be
induced from a real quantum state, given a particular quantum IVPM,
i.e., using a particular measurement apparatus, our investigation
leaves an open question of whether behind it there exists a ``real''
state. There are several possibilities.

One possibility is to trace back the original definition of a quantum
state. A quantum state could be formulated as a density matrix in
von Neumann's formalism~\cite{vonNeumann1955,Varadarajan2008}. Later,
Mackey defined a quantum state as additive probability measure as
we did in definition~\ref{def:QuantumProbabilitySpace}~\cite{inun.606849520040101,Varadarajan2008}.
Although Gleason proved Mackey states are von Neumann states in dimension
$\ge3$, it might not have to be when we generalized Mackey states
to quantum IVPMs, and maybe we should use quantum IVPMs as the most
primitive definition of states.

The other possibility focus on the meaning of the probability. Since
a quantum IVPM cannot be defined without consider a particular measurement
apparatus, and the information extracted from this apparatus, the
interval-valued probability itself could be considered subjective,
and could be interpreted in the Bayesian sense as ``lack of knowledge''.
This, and similar arguments elegantly expressed by recent work on
Quantum Bayesianism~\cite{Fuchs2010,VonBaeyer2016,Fuchs2012}, which
suggests that the quantum state is more like an interactive system
in computer science parlance. In other words, the state is only determined
by its interactions with the environment or observers. A rough but
useful analogy in the context of computer science would be a web server:
the state of such a system cannot be said to exist independently of
the interactions with clients; each interaction with a client determines
a future state of the system and may affect future interactions. Crucially,
each client may have a \emph{different} view of the web server that
is consistent with their own history of interactions. This computational
perspective agrees with the philosophical position of QBism~\cite{VonBaeyer2016}
that asserts that the quantum state is subjective: each observer has
a different view of the quantum system that is consistent with their
previous observations and that allows that observer, independently
of other observers, to assign beliefs (i.e., probabilities) to possible
future interactions.
